Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-10-17 paper-bedpca.tex
save UKBB ancestry

\subsection{Outlier detection in PCA [SPLIT IN MULTIPLE SECTIONS]}

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use the same procedure as described by \cite{prive2017efficient}: we compute pcadapt statistics \cite[]{luu2017pcadapt} that summarises the contribution of each variant in all PC loadings computed. This captures both population-differentiating variants as well as long-range LD regions. We then smooth the statistics [BE MORE PRECISE?] to mostly capture consecutive outliers that corresponds to long-range LD regions.

@@

We use the 1000 Genomes (1000G) Project data (section \ref{1000G}) to infer the ancestry of individuals from a study dataset following the following steps: 1) we compute the robust center and covariance parameters of the Mahalanobis distance (section \ref{maha}) for each of the 26 populations of the 10000G data; 2) we project individuals from the study dataset onto the PCA space computed using the 1000G data (section \ref{proj}); 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we affect each individual to the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e. individuals far from every population of the 1000G data).

@@

@@

[LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]

@@

@@

% latex table generated in R 3.5.2 by xtable 1.8-4 package
% Wed Jun 19 15:19:38 2019
\begin{table}[!htpb]
\caption{\label{tab:AUC-simu2}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
Scenario & stdCT & maxCT & SCT & lassosum \\ 
\hline
100  & 77.4 [76.0-78.8] & 83.9 [83.4-84.4] & 83.1 [82.6-83.6] & 80.1 [79.5-80.8] \\ 
10K  & 69.4 [68.4-70.5] & 73.0 [72.5-73.4] & 72.9 [72.5-73.3] & 71.2 [70.6-71.7] \\ 
1M   & 64.0 [63.6-64.4] & 64.0 [63.6-64.4] & 62.7 [62.3-63.0] & 64.1 [63.3-64.8] \\ 
2chr & 70.0 [68.8-71.2] & 74.4 [73.6-75.2] & 78.5 [77.9-79.1] & 73.2 [72.5-73.8] \\ 
err  & 67.0 [66.0-68.1] & 68.6 [67.7-69.5] & 69.5 [68.9-70.1] & 65.6 [64.9-66.3] \\ 
HLA  & 74.8 [72.9-76.3] & 75.3 [73.5-76.9] & 76.4 [74.5-78.0] & 75.8 [74.2-77.2] \\ 
\hline
\end{tabular}
\end{table}

>>>>

\subsection{Detecting LD structure in PCA}

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a slightly modified version of the procedure described by \cite{prive2017efficient}. 
To summarise the contribution of each variant in all PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of PC loadings.  
This captures both population-differentiating variants as well as long-range LD regions. 
Then, to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.

\subsection{Detecting outlier samples in PCA}

@@

@@

\subsection{Ancestry estimation\label{ancestry}}

We use the 1000 Genomes (1000G) Project data (section \ref{1000G}) to infer the ancestry of individuals from a study dataset following the following steps: 1) we compute the robust center and covariance parameters of the Mahalanobis distance (section \ref{maha}) for each of the 26 populations of the 10000G data; 2) we project individuals from the study dataset onto the PCA space computed using the 1000G data (section \ref{proj}); 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we affect each individual to the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.\ individuals far from every population of the 1000G data).

@@

[REDO WITH OUTLIER REMOVAL]

We performed ancestry estimation of the individuals in the UK Biobank using the 1000G data (Section \ref{ancestry}), discarding individuals with unknown ancestry or mixed ancestry. 
We did not infer ancestry for 15.5\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.4\%, 10.5\% and 19.8\%, while this represented between 48.5\% (``Chinese'') and 81.4\% (``Bangladeshi''). 
In other words, most of the UK Biobank individual who self reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes some ``Bengali in Bangladesh'' population.
Only 10 people out of 401,048 were wrongly classified in super population of the 1000G; e.g.\ one Chinese was classified as European by our method (Table \ref{tab:ancestry-pred}). 
Our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Thu Oct 17 13:58:41 2019
\begin{table}[ht]
\centering
\caption{Self-reported ancestry (top) of UKBB individuals and their matching to 1000G populations (left) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|}
  \hline
 & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black \\ 
  \hline
AFR\_ACB &  &  &  &  &  &  &  &  &  & 1635 & 510 & 27 \\ 
  AFR\_ASW &  &  &  &  &  &  &  &  &  & 488 & 23 & 14 \\ 
  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 103 &  \\ 
  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 15 &  \\ 
  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 38 &  \\ 
  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 25 &  \\ 
  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 433 & 4 \\ 
   \hline
AMR\_CLM & 1 &  &  & 53 &  &  &  &  &  &  &  &  \\ 
  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  \\ 
  AMR\_PEL &  &  &  & 1 &  &  &  &  &  &  &  &  \\ 
  AMR\_PUR &  &  &  &  &  &  &  &  &  &  &  &  \\ 
   \hline
EAS\_CDX &  &  &  &  &  &  &  & 1 &  &  &  &  \\ 
  EAS\_CHB &  &  & 1 &  &  &  &  & 408 & 11 &  &  &  \\ 
  EAS\_CHS &  &  &  &  &  &  &  & 350 & 6 &  &  &  \\ 
  EAS\_JPT &  &  &  &  &  &  &  & 2 & 24 &  &  &  \\ 
  EAS\_KHV &  &  &  &  &  &  &  & 3 & 16 &  &  &  \\ 
   \hline
EUR\_CEU & 293299 & 5483 & 327 & 4486 &  &  &  & 1 &  &  &  &  \\ 
  EUR\_FIN & 2 &  &  & 116 &  &  &  &  &  &  &  &  \\ 
  EUR\_GBR & 91862 & 4881 & 99 & 765 &  &  &  &  &  &  & 1 &  \\ 
  EUR\_IBS & 32 &  & 4 & 481 &  &  &  &  &  &  &  &  \\ 
  EUR\_TSI & 20 &  & 1 & 166 &  &  &  &  &  &  &  &  \\ 
   \hline
SAS\_BEB &  &  &  &  & 12 &  & 50 &  &  &  &  &  \\ 
  SAS\_GIH &  &  &  &  & 302 &  &  &  &  &  &  &  \\ 
  SAS\_ITU & 1 &  &  &  & 307 & 3 &  &  & 251 & 3 &  &  \\ 
  SAS\_PJL &  &  &  &  & 1452 & 712 &  &  & 82 & 1 &  &  \\ 
  SAS\_STU &  &  &  &  & 42 &  &  &  & 172 &  &  &  \\ 
   \hline
\end{tabular}
\end{table}

\noindent[OKAY TO CLASSIFY SOME OTHER WHITE AS CENTRAL/SOUTH AMERICANS?]

\noindent[INCLUDE RESULTS FOR SIMPLE PROJ? -> HIST PVAL BAD -> 38\% NOT MATCHED, ESP PAKISTANI]

@@

\noindent[LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]

@@

\subsection*{Supplementary figures}

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.8\textwidth]{hist-pval-max.pdf}}
\caption{Maximum p-values based on robust Mahalabonis distances of UK Biobank individuals from each of the 26 1000G populations.
\label{fig:ancestry-pval}}
\end{figure}
